\normalfont\spacedallcaps{The economic consequences of different attitudes of a policy maker: a combined epidemiological-econometric study}

### 2021-03-17 Manuscript/manuscript.tex
PUb

known as the basic reproduction factor. This number is the average number of people that are infected by a single individual and describes if the epidemic is in a negative feedback (R$_{0}$<1),  stationary (R$_{0}$=1),  or in a positive feedback (R$_{0}$>1).  As consequence if the epidemic is in a negative feedback will be dissipate, while if it is in positive feedback will grow.  Note that in this simple model, since the parameters are not time dependent this factor is constant.  As performed by Ferrari et al. \cite{ferrari2020modelling},  when time dependent parameters $\beta$ and $\gamma$ are taken in to account the reproduction factor R$_{0}$ become time depended: thus scholar rename it as $R_{t}$.  For the present work,  we limited to constant parameters: in particular we considered the parameter estimation for Lombardy provided by Neves and Guerrero in Ref \cite{neves2020predicting}: $\beta$ was set equal to 0.55 while $\gamma$ equal to $\frac{1}{7}$.  The $\mu$ was set in order to keep in to account the calculated lethality for Italy: 1 \% \cite{fernandez2020estimating} (see also the very fine analysis made by the Institute for International Political Studies (ISPI) \cite{ISPI})\footnote{The authors are aware that this ratio is far from being homogeneus for the the different ages of population: however if this factor is taken into account,  it requires the solution of a system of partial derivative equations. In this case the numerical calculations become much more complicated}.  The overall population $N$ was set to 60M,  in order to simulate Italian population.  Within the daily temporal evolution of this model,  that was obtained by numerically solving the differential equation above via the \textit{DeSolve} package,  we considered a trigger activated by the PM every 7 days: if the number of infected people normalized by the overall population is more than $1\times 10^{-7}$,  there is a probability that the PM impose laws that multiply the $\beta$ factor by $0.7$,  if the normalized infected people are more than 10 the previous threshold he/she will impose with a certain probability restrictions that multiply $\beta$ starting $\beta$ factor by $0.25$;  finally,  if the  threshold is exceeded more than 50 times, the PM will impose with a certain probability restrictive measures that multiply $\beta$ by $0.025$.  These attenuation parameters were adjusted taking in consideration results of Marziano et al. in Ref. \cite{marziano2021retrospective}.  Therefore such trigger make the  R$_{0}$ parameter time dependent,  although in to a discrete way.  As we said the PM acts with a certain probability,  more formally stochastically: each week a random number (from zero to one) is extracted: if this is higher than a certain threshold,  the relative restrictive decision is taken, otherwise not. The threshold value captures the PM attitude to impose the lock-down: lower values model a careful PM,  hight value a lazy one.  In this way the model is able to simulate different scenarios for the different PM attitude: as we will see this can produce two very different results.  At the end of the simulation,  beside the values given by the standard SIRD model (recovered and deaths),  the algorithm also provide the number of weeks in which each restriction was active: these values are then used for the economic model in order to evaluate the economic effect due to the restrictions and PM strategy.  It is worth nothing that here as lock-down we considered only the national one that was applied in the first wave of the epidemics: from March 2020 to June 2020. This choice is motivated by the fact that modelling a unique system is more simpler with respect to modelling an ensemble of a communicating clusters that represent regions or provinces: therefore,  if one is interested to model the second wave of pandemics such approach should be undertaken.  Moreover in this latter case,  as a further degree of complexity,  the economic data described in the next subsection must be at regional or province level,  and as far as the authors know,  such data are not available.  For these reasons here the modelling will be always referred to national data and national lock-downs,  therefore the period after June when starting from October regional lock-down were imposed,  will be not considered.  Another point that is it is worth mentioning,  is that in this model, on the contrary with respect to the SIS one, individual can not re-infect: concerning the COVID-19 the possibility of reinfection is still discussed among scholars \cite{ledford2020covid,iwasaki2021reinfections}.  Basically to the know of author it is established that the immunity last at least 8 months  \cite{dan2021immunological} and very few cases of reinfection are reported.  Thus the immunity considered in the model,  for an interval of one year seems almost  a realistic approximation.

@@

Now that we have the parameters for the economic impact given by the lock-downs,  we are ready to discuss the simulations that output the epidemiological and economic consequences of the attitude of PM.  In Tab.  \ref{Table_Epidemiological_effect} and \ref{Table_Epidemiological_cost}  the outcomes of the epidemiological model are given,  while the SIRD curves for the scenarios are provided in Fig.  \ref{SIRD_scenario}; on the other side the effect on selling and subsidies for both scenarios are given in Fig. \ref{Selling_sector_effects} and \ref{CIG_sector_effects}.  In order to assure the \textit{ceteris pariubus} condition we used for both scenario the same set of random numbers.  First,  we see that where the PM is more reactive we have small different waves of epidemics,  while if the PM is poorly reactive only one intense wave is present,  indeed in the latter scenario the laziness of PM to apply the lock-down in the first weeks produce the full infection peak of the standard SIRD model: when the PM acts is too late,  as shown in Fig.  \ref{SIRD_scenario}, since the pandemics has almost hit the large part of the population.  This can be pointed out in Fig. \ref{R_t_scenarios} by noting that that the reproduction factor remains at its maximum level for large number of weeks after the beginning of the pandemic. This also explains why in the second half of the scenario the lock-downs are so week: the remaining part of the population is immune.,  an thus no further action is required.  On the contrary in the first scenario the PM is able,  using the lock-downs to transform the SIRD peak in to small waves (Fig.  \ref{SIRD_scenario}).  As shown in Fig.  \ref{R_t_scenarios} this result is basically obtained by keeping the reproduction factor in the value of medium lock-down value.  These two theoretical model correspond,  in practice,  for the lazy PM to the one considered by the PMs that aimed to heard immunity,  while,  concerning the active-PM,  to the PMs that would not saturate the saturate the health system and aimed to reduce the death at the minimum.  It is interesting to point out that the change between the scenarios develops in a discontinuous way as showed in Fig.  \ref{Phase_transition},   as the reactivity of the PM is changed,  we can consider this as a phase transition of a physical system (it is interesting to point out the closness with respect to the gold-standard diagram of Ising model \cite{Ising,Ising_2}). This explains our choice to consider only two scenarios: indeed we considered only a sample for each phase.  It is worth nothing that a similar result,  within a different epidemiological model,  was obtained by Balcan and Vespignani in Ref.  \cite{Vespignani}.  The explanation of this similar behaviour is that both model have a stochastic component,  that as pointed out by Balcan and Vespignani,  give the phase transition: indeed while in their model this was directly related to the contagion probability,  here this is indirectly made stochastically by the decision of PM that modify the $\beta$ parameter and so the transmission rate.  Moving to the socio-economic outcome of the model used we see, from Tab. \ref{Table_Epidemiological_effect} and \ref{Table_Epidemiological_cost},  that in the small wave scenario,  the attitude of the PM largely reduce the overall number of deaths and recovered (and so the cumulative cost for taking of patients) by a consistent use of the confinement.  This has an  immediate drawback on the economic data: as shown in Fig. \ref{Selling_sector_effects} and \ref{CIG_sector_effects}, the loses for selling and the use of subsidies are widely large with respect to the one-wave scenario.  In principle one can be attempted to find the PM attitude that minimize the overall cost (deaths,  infected,  selling and subsidies): in the authors view, this scenario is not realistic since,  actually,  the Italian (but also many other European) criminal law does not allow this option: although there is an economic cost for a life in terms of compensation,  the actual criminal law does not consider an amount of money comparable to a money sum,  on contrary it gives a value ex post not ex ante.  For this reason here we considered the epidemiological consequences separated with respect to the economic ones. A more intriguing issue is come up when the economic cost is causally associated to an number indirect of deaths (for instance if people does not have the money for food or other first necessity goods ),  in this case the two factor (epidemiological and economic) can be,  in principle,  summed.  We say in principle because the jurisprudence is here very reduced or totally missing since the pandemics are rare events.  For the present study these indirect deaths are,  by now,  not easily to quantify and thus we did not considered this option.

>>>>

known as the basic reproduction factor. This number is the average number of people that are infected by a single individual and describes if the epidemic is in a negative feedback (R$_{0}$<1),  stationary (R$_{0}$=1),  or in a positive feedback (R$_{0}$>1).  As consequence if the epidemic is in a negative feedback will be dissipate, while if it is in positive feedback will grow.  Note that in this simple model, since the parameters are not time dependent this factor is constant.  As performed by Ferrari et al. \cite{ferrari2020modelling},  when time dependent parameters $\beta$ and $\gamma$ are taken in to account the reproduction factor R$_{0}$ become time depended: thus scholar rename it as $R_{t}$.  For the present work,  we limited to constant parameters: in particular we considered the parameter estimation for Lombardy provided by Neves and Guerrero in Ref \cite{neves2020predicting}: $\beta$ was set equal to 0.55 while $\gamma$ equal to $\frac{1}{7}$.  The $\mu$ was set in order to keep in to account the calculated lethality for Italy: 1 \% \cite{fernandez2020estimating} (see also the very fine analysis made by the Institute for International Political Studies (ISPI) \cite{ISPI})\footnote{The authors are aware that this ratio is far from being homogeneus for the the different ages of population: however if this factor is taken into account,  it requires the solution of a system of partial derivative equations. In this case the numerical calculations become much more complicated}.  The overall population $N$ was set to 60M,  in order to simulate Italian population.  Within the daily temporal evolution of this model,  that was obtained by numerically solving the differential equation above via the \textit{DeSolve} package, we considered a trigger activated by the PM every 7 days: if the number of infected people normalized by the overall population is more than $1\times 10^{-7}$,  there is a probability that the PM imposes laws that reduces the $\beta$ factor by a multiplicative factor equal to $0.7$ (and thus the reproduction factor $R_{0}$), if the normalized infected people are more than 10 the previous threshold he/she will impose with a certain probability restrictions that reduce the $\beta$ factor by a multiplicative factor of $0.25$; finally, if the threshold is exceeded more than 50 times, the PM will impose with a certain probability restrictive measures that reduce the $\beta$ a multiplicative factor equal to $0.025$. These attenuation parameters were adjusted taking in consideration results of Marziano et al. in Ref. \cite{marziano2021retrospective}.  Therefore such trigger make the  R$_{0}$ parameter time dependent,  although in to a discrete way.  As we said the PM acts with a certain probability,  more formally stochastically; each week a random number (from zero to one) is extracted: if this is higher than a certain threshold,  the relative restrictive decision is taken, otherwise not. The threshold value captures the PM attitude to impose the lock-down: lower values model a careful PM, hight value a lazy one. In this way the model is able to simulate different scenarios for the different PM attitude: as we will see this can produce two very different results.  At the end of the simulation,  beside the values given by the standard SIRD model (recovered and deaths),  the algorithm also provide the number of weeks in which each restriction was active: these values are then used for the economic model in order to evaluate the economic effect due to the restrictions and PM strategy. It is worth nothing that here as lock-down we considered only the national one that was applied in the first wave of the epidemics: from March 2020 to June 2020. This choice is motivated by the fact that modelling a unique system is easier with respect to modelling an ensemble of a communicating clusters that represent regions or provinces: therefore, if one is interested to model the second wave of pandemics such approach should be undertaken. Moreover in this latter case,  as a further degree of complexity,  the economic data described in the next subsection must be at regional or province level,  and as far as the authors know,such data are not available.  For these reasons here the modelling will be always referred to national data and national lock-downs,  therefore the period after June when starting from October regional lock-down were imposed,  will be not considered.  Another point that is it is worth mentioning,  is that in this model, on the contrary with respect to the SIS one, individual can not re-infect: concerning the COVID-19 the possibility of reinfection is still discussed among scholars \cite{ledford2020covid,iwasaki2021reinfections}.  Basically to the know of author it is established that the immunity last at least 8 months  \cite{dan2021immunological} and very few cases of reinfection are reported.  Thus the immunity considered in the model,  for an interval of one year seems almost  a realistic approximation.

@@

Now that we have the parameters for the economic impact given by the lock-downs,  we are ready to discuss the simulations that output the epidemiological and economic consequences of the attitude of PM.  In Tab.  \ref{Table_Epidemiological_effect} and \ref{Table_Epidemiological_cost}  the outcomes of the epidemiological model are given,  while the SIRD curves for the scenarios are provided in Fig.  \ref{SIRD_scenario}; on the other side the effect on selling and subsidies for both scenarios are given in Fig. \ref{Selling_sector_effects} and \ref{CIG_sector_effects}.  In order to assure the \textit{ceteris pariubus} condition we used for both scenario the same set of random numbers.  First,  we see that where the PM is more reactive we have small different waves of epidemics,  while if the PM is poorly reactive only one intense wave is present,  indeed in the latter scenario the laziness of PM to apply the lock-down in the first weeks produce the full infection peak of the standard SIRD model: when the PM acts is too late,  as shown in Fig.  \ref{SIRD_scenario}, since the pandemics has almost hit the large part of the population.  This can be pointed out in Fig. \ref{R_t_scenarios} by noting that that the reproduction factor remains at its maximum level for large number of weeks after the beginning of the pandemic. This also explains why in the second half of the scenario the lock-downs are so week: the remaining part of the population is immune.,  an thus no further action is required.  On the contrary in the first scenario the PM is able,  using the lock-downs to transform the SIRD peak in to small waves (Fig.  \ref{SIRD_scenario}).  As shown in Fig.  \ref{R_t_scenarios} this result is basically obtained by keeping the reproduction factor in the value of medium lock-down value.  These two theoretical model correspond,  in practice,  for the lazy PM to the one considered by the PMs that aimed to heard immunity,  while,  concerning the active-PM,  to the PMs that would not saturate the saturate the health system and aimed to reduce the death at the minimum.  It is interesting to point out that the change between the scenarios develops in a discontinuous way as showed in Fig.  \ref{Phase_transition}, as the reactivity of the PM is changed, we can consider this as a phase transition of a physical system (e.g. consider the gold-standard diagram of Ising model \cite{Ising,Ising_2}). This explains our choice to consider only two scenarios: indeed we considered only a sample for each phase.  It is worth nothing that a similar result,  within a different epidemiological model,  was obtained by Balcan and Vespignani in Ref.  \cite{Vespignani}.  The explanation of this similar behaviour is that both model have a stochastic component,  that as pointed out by Balcan and Vespignani,  give the phase transition: indeed while in their model this was directly related to the contagion probability,  here this is indirectly made stochastically by the decision of PM that modify the $\beta$ parameter and so the transmission rate.  Moving to the socio-economic outcome of the model used we see, from Tab. \ref{Table_Epidemiological_effect} and \ref{Table_Epidemiological_cost},  that in the small wave scenario,  the attitude of the PM largely reduce the overall number of deaths and recovered (and so the cumulative cost for taking of patients) by a consistent use of the confinement.  This has an  immediate drawback on the economic data: as shown in Fig. \ref{Selling_sector_effects} and \ref{CIG_sector_effects}, the loses for selling and the use of subsidies are widely large with respect to the one-wave scenario.  In principle one can be attempted to find the PM attitude that minimize the overall cost (deaths,  infected,  selling and subsidies): in the authors view, this scenario is not realistic since,  actually,  the Italian (but also many other European) criminal law does not allow this option: although there is an economic cost for a life in terms of compensation,  the actual criminal law does not consider an amount of money comparable to a money sum,  on contrary it gives a value ex post not ex ante.  For this reason here we considered the epidemiological consequences separated with respect to the economic ones. A more intriguing issue is come up when the economic cost is causally associated to an number indirect of deaths (for instance if people does not have the money for food or other first necessity goods ),  in this case the two factor (epidemiological and economic) can be,  in principle,  summed.  We say in principle because the jurisprudence is here very reduced or totally missing since the pandemics are rare events.  For the present study these indirect deaths are,  by now,  not easily to quantify and thus we did not considered this option.
